Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Tumors

### 2018-06-04 submission/manuscript.Rmd
smooth over the methods

**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment for their tumor (n=`r length(metaf$initial)`). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days. Our use of treatment has been previously defined as encompassing removal of a tumor with or without chemotherapy and radiation [@normalization_sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Decleration.

**Measuring specific SCFAs.** Our protocol for the measurement of acetate, butyrate, and propionate followed a previously published protocol [@scfa_measures_venkataraman2016]. The following changes to this protocol included the use of frozen fecal samples suspended in 1ml of PBS instead of fecal suspensions in DNA Genotek OmniGut tubes, and the use of the acutal weight of fecal samples instead of the average weight for SCFA concentration normalizations. These changes did not affect the overall median concentrations of these SCFAs between the two studies (see Table 1 [@scfa_measures_venkataraman2016] and Figure 1 in this report).

**16s rRNA gene sequencing.** The workflow and processing have been described previously [@Schloss2009; @Kozich2013; @normalization_sze2017]. The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity used the OptClust algorithm [@opticlust_westcott2017].

**Generating imputed metagenomes.** The use of PICRUSt version 1.1.2 with the recommended standard operating protocol [@Langille2013] was used. Briefly, the mothur shared file and metadata was converted into a biom formated table using the biom convert function, the subsequent biom file was processed with the normalize_by_copy_number.py function, and subsequent imputed metagenomes created using the predict_metagenomes.py function.

**Obtaining OPFs from metagenomes.** A subset of the cross-sectional group (n=`r length(metai$sample)`) containing a total of `r length(geof_samples$X1)` individuals (normal n=`r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()` normal, adenoma n=`r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()`, and carcinoma n=`r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()`) was shotgun sequenced on an Illumina HiSeq using 125 bp paired end reads and a previously described method [@Hannigan2017]. Briefly, the sequences were quality filtered and sequences aligning to the human genome were removed prior to contig assembly with MEGAHIT [@Li2015]. Open Reading Frames (ORFs) were identified using Prodigal [@Hyatt2010], counts generated using Diamond [@Buchfink2014], subsequent clustering into Operational Protein Families (OPFs) used mmseq2 [@Steinegger2017], and OPF alignment used the KEGG database [@Kanehisa2015].

@@

**Random Forest Models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for both classification and regression models [@randomforest_citation_2002]. This was repeated on 100 differen 80/20 splits of the data to generate a reasonable range for the AUC of the model. The reported AUCs, unless otherwise specified, are for the test sets. Classification models were built to group normal versus adenoma and normal versus carcinoma or high versus low SCFA concentrations.

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, differences between the different disease groups used a Kruskal-Walis rank sum test with a Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995]. First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma). We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment. Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumor using Random Forest models. Next, the imputed gene counts of important mediators of SCFA creation were tested. Finally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed. Finally, models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using the Random Forest algorithm.

>>>>

**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment of a patient's tumor (n=`r length(metaf$initial)`). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days. Our use of treatment has been previously defined as encompassing removal of a tumor with or without chemotherapy and radiation [@normalization_sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Decleration.

**Measuring specific SCFAs.** Our protocol for the measurement of acetate, butyrate, and propionate followed a previously published protocol that used a High-Performance Liquid Chromatography (HPLC) machine [@scfa_measures_venkataraman2016]. The following changes to this protocol included the use of frozen fecal samples suspended in 1ml of PBS instead of fecal suspensions in DNA Genotek OmniGut tubes, and the use of the acutal weight of fecal samples instead of the average weight for SCFA concentration normalizations. These methodological changes did not affect the overall median concentrations of these SCFAs between the two studies (see Table 1 [@scfa_measures_venkataraman2016] and Figure 1 here).

**16s rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @normalization_sze2017]. The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptClust algorithm [@opticlust_westcott2017].

**Generating imputed metagenomes.** The use of PICRUSt version 1.1.2 with the recommended standard operating protocol [@Langille2013] was used. Briefly, the mothur shared file and metadata was converted into a biom formated table using the biom convert function, the subsequent biom file was processed with the 'normalize_by_copy_number.py' function, and subsequent imputed metagenomes created using the 'predict_metagenomes.py' function.

**Obtaining Operational Protein Families from metagenomes.** A subset of the cross-sectional group (n=`r length(metai$sample)`) containing a total of `r length(geof_samples$X1)` individuals (normal n=`r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()`, adenoma n=`r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()`, and carcinoma n=`r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()`) was shotgun sequenced on an Illumina HiSeq using 125 bp paired end reads and a previously described method [@Hannigan2017]. Briefly, the sequences were quality filtered and sequences aligning to the human genome were removed prior to contig assembly with MEGAHIT [@Li2015]. Open Reading Frames (ORFs) were identified using Prodigal [@Hyatt2010], counts generated using Diamond [@Buchfink2014], subsequent clustering into Operational Protein Families (OPFs) used mmseq2 [@Steinegger2017], and OPF alignment used the KEGG database [@Kanehisa2015].

@@

**Random Forest models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for classification models [@randomforest_citation_2002]. This was repeated on 100 different 80/20 splits of the data to generate a reasonable range for the AUC of the model. The reported AUCs, unless otherwise specified, are for the test sets. The classification models were built to group normal versus adenoma, normal versus carcinoma, and high versus low SCFA concentrations.

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, a Kruskal-Walis rank sum test with a Dunn's post-hoc test was used to assess differences between the groups used. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995]. First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma). We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment. Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumor using Random Forest models. Next, the imputed gene counts of important mediators of SCFA synthesis was tested. Additionally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed. Finally, models to classify high or low SCFA concentration based on the median of each SCFA and 16S rRNA gene sequencing data was created using the Random Forest algorithm.

### 2018-06-04 submission/manuscript.tex
smooth over the methods

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (26, 27). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment for their tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (26).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (26, 27). The University
of Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Decleration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol (28). The following changes to this protocol included the use
of frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the acutal
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These changes did not affect the overall
median concentrations of these SCFAs between the two studies (see Table
1 (28) and Figure 1 in this report).

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
described previously (26, 29, 30). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity used the OptClust algorithm (31).

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (20) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formated table using the biom convert function, the subsequent biom file
was processed with the normalize\_by\_copy\_number.py function, and
subsequent imputed metagenomes created using the predict\_metagenomes.py
function.

\textbf{Obtaining OPFs from metagenomes.} A subset of the
cross-sectional group (n=490) containing a total of 85 individuals
(normal n=29 normal, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (32). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (33). Open Reading Frames (ORFs) were
identified using Prodigal (34), counts generated using Diamond (35),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (36), and OPF alignment used the KEGG database (37).

@@

\textbf{Random Forest Models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for both classification and regression models
(19). This was repeated on 100 differen 80/20 splits of the data to
generate a reasonable range for the AUC of the model. The reported AUCs,
unless otherwise specified, are for the test sets. Classification models
were built to group normal versus adenoma and normal versus carcinoma or
high versus low SCFA concentrations.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (39). Generally, differences between the
different disease groups used a Kruskal-Walis rank sum test with a
Dunn's post-hoc test. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (40). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
alone or OTUs and SCFAs were better able to classify individuals with
and without tumor using Random Forest models. Next, the imputed gene
counts of important mediators of SCFA creation were tested. Finally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed. Finally, models to classify high or low
SCFA concentration based on 16S rRNA gene sequencing data were created
using the Random Forest algorithm.

>>>>

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (26, 27). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment of a patient's tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (26).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (26, 27). The University
of Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Decleration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol that used a High-Performance Liquid Chromatography (HPLC)
machine (28). The following changes to this protocol included the use of
frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the acutal
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These methodological changes did not
affect the overall median concentrations of these SCFAs between the two
studies (see Table 1 (28) and Figure 1 here).

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
previously described (26, 29, 30). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity using the OptClust algorithm (31).

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (20) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formated table using the biom convert function, the subsequent biom file
was processed with the `normalize\_by\_copy\_number.py' function, and
subsequent imputed metagenomes created using the
`predict\_metagenomes.py' function.

\textbf{Obtaining Operational Protein Families from metagenomes.} A
subset of the cross-sectional group (n=490) containing a total of 85
individuals (normal n=29, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (32). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (33). Open Reading Frames (ORFs) were
identified using Prodigal (34), counts generated using Diamond (35),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (36), and OPF alignment used the KEGG database (37).

@@

\textbf{Random Forest models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for classification models (19). This was
repeated on 100 different 80/20 splits of the data to generate a
reasonable range for the AUC of the model. The reported AUCs, unless
otherwise specified, are for the test sets. The classification models
were built to group normal versus adenoma, normal versus carcinoma, and
high versus low SCFA concentrations.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (39). Generally, a Kruskal-Walis rank sum
test with a Dunn's post-hoc test was used to assess differences between
the groups used. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (40). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
alone or OTUs and SCFAs were better able to classify individuals with
and without tumor using Random Forest models. Next, the imputed gene
counts of important mediators of SCFA synthesis was tested.
Additionally, the counts generated for OPFs that matched important genes
involved with SCFA creation were analyzed. Finally, models to classify
high or low SCFA concentration based on the median of each SCFA and 16S
rRNA gene sequencing data was created using the Random Forest algorithm.
